quollr: An R Package for Visualizing 2D Models from Nonlinear Dimension Reductions in High Dimensional Space

### 2024-04-02 paper-quollr.Rmd
Added dependency info

---
title: "quollr: An R Package for Visualizing 2D Models from Nonlinear Dimension Reductions in High Dimensional Space"
abstract: >
  Non-Linear Dimension Reduction (NLDR) techniques have emerged as powerful tools to visualize high-dimensional ($p$-D) data. However, their complexity and parameter choices may lead to distrustful or misleading results. The R package \CRANpkg{quollr} is developed as a new tool to help to determine which method, which parameter choice provide the most useful representation of high-D data. Clustering data from \CRANpkg{cardinalR} package, is used to illustrate the algorithm and its use within the paackage.
draft: true
author:  
  - name: Jayani P.G. Lakshika
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    url: https://jayanilakshika.netlify.app/
    orcid: 0000-0002-6265-6481
    email:  \email{jayani.piyadigamage@monash.edu}
  - name: Dianne Cook
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    url: http://www.dicook.org/
    email: dicook@monash.edu
    orcid: 0000-0002-3813-7155
  - name: Paul Harrison
    affiliation: Monash University
    address: MGBP, BDInstitute, VIC 3800 Australia
    email: paul.harrison@monash.edu
    orcid: 0000-0002-3980-268X
  - name: Michael Lydeamore
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    email: michael.lydeamore@monash.edu
    orcid: 0000-0001-6515-827X
  - name: Thiyanga S. Talagala
    affiliation: University of Sri Jayewardenepura
    address: Department of Statistics, Gangodawila, Nugegoda 10100 Sri Lanka
    url: https://thiyanga.netlify.app/
    email: ttalagala@sjp.ac.lk
    orcid: 0000-0002-0656-9789
type: package
creative_commons: CC BY
date: "`r Sys.Date()`"
preamble: >
  \usepackage{amsmath}
  \usepackage{array}
output: 
 rjtools::rjournal_web_article:
    css: "style.css"
    keep_md: true
bibliography: RJreferences.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, 
  cache=FALSE, 
  message=FALSE, 
  warning=FALSE)

@@

```{r load-libraries}
library(quollr)
library(tibble)
library(knitr)
library(kableExtra)
library(ggplot2)
library(dplyr)
library(ggbeeswarm)
library(patchwork)
library(Rtsne)

@@

@@

>>>>

---
title: "quollr: An R Package for Visualizing 2D Models from Nonlinear Dimension Reductions in High Dimensional Space"
abstract: >
  Non-Linear Dimension Reduction (NLDR) techniques have emerged as powerful tools to visualize high-dimensional ($p-D$) data in low-diemnsioanl space. However, their complexity and hyperparameter choices may lead to distrustful or misleading results. The R package \CRANpkg{quollr} is developed as a new tool to help to determine which method, which hyperparameter choice provide the most accurate representation of $p-D$ data. Clustering data from \CRANpkg{cardinalR} package, is used to illustrate the algorithm and its use within the package.
draft: true
author:  
  - name: Jayani P.G. Lakshika
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    url: https://jayanilakshika.netlify.app/
    orcid: 0000-0002-6265-6481
    email:  \email{jayani.piyadigamage@monash.edu}
  - name: Dianne Cook
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    url: http://www.dicook.org/
    email: dicook@monash.edu
    orcid: 0000-0002-3813-7155
  - name: Paul Harrison
    affiliation: Monash University
    address: MGBP, BDInstitute, VIC 3800 Australia
    email: paul.harrison@monash.edu
    orcid: 0000-0002-3980-268X
  - name: Michael Lydeamore
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    email: michael.lydeamore@monash.edu
    orcid: 0000-0001-6515-827X
  - name: Thiyanga S. Talagala
    affiliation: University of Sri Jayewardenepura
    address: Department of Statistics, Gangodawila, Nugegoda 10100 Sri Lanka
    url: https://thiyanga.netlify.app/
    email: ttalagala@sjp.ac.lk
    orcid: 0000-0002-0656-9789
type: package
creative_commons: CC BY
date: "`r Sys.Date()`"
preamble: >
  \usepackage{amsmath}
  \usepackage{array}
output: 
 rjtools::rjournal_web_article:
    css: "style.css"
    keep_md: true
bibliography: RJreferences.bib
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
options(repos = "https://cloud.r-project.org") ## set up CRAN mirror
knitr::opts_chunk$set(
  echo = FALSE, 
  cache=FALSE, 
  message=FALSE, 
  warning=FALSE,
  out.width = "100%")

@@

```{r load-libraries}
library(quollr)
library(tibble)
library(knitr)
library(kableExtra)
library(ggplot2)
library(dplyr)
library(ggbeeswarm)
library(patchwork)
library(Rtsne)
library(readr)
library(tools)

@@

```{r}
tSNE_data <- read_rds("data/s_curve/s_curve_tsne_27.rds")

plot_list1 <- tSNE_data |>
  ggplot(aes(x = tSNE1,
             y = tSNE2)) +
  geom_point(alpha=0.1, size=1, colour="#e41a1c") +
  annotate(geom = 'text', label = 'a', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

```{r}
UMAP_data <- read_rds(file = "data/s_curve/s_curve_umap.rds")

#(n-neighbors: 50)
plot_list2 <- UMAP_data |>
  ggplot(aes(x = UMAP1,
             y = UMAP2)) +
  geom_point(alpha=0.1, size=1, colour="#377eb8") +
  annotate(geom = 'text', label = 'b', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

```{r}
#(knn: 5)
PHATE_data <- read_rds(file = "data/s_curve/s_curve_phate.rds")

plot_list3 <- PHATE_data |>
  ggplot(aes(x = PHATE1,
             y = PHATE2)) +
  geom_point(alpha=0.1, size=1, colour="#4daf4a") +
  annotate(geom = 'text', label = 'c', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

```{r}
TriMAP_data <- read_rds(file = "data/s_curve/s_curve_trimap.rds")

#(n-inliers: 5, \n n-outliers: 4, n-random: 3)
plot_list4 <- TriMAP_data |>
  ggplot(aes(x = TriMAP1,
             y = TriMAP2)) +
  geom_point(alpha=0.1, size=1, colour="#984ea3") +
  annotate(geom = 'text', label = 'd', x = Inf, y = Inf, hjust = 1.5, vjust = 1.5, size = 3)

```{r}
PaCMAP_data <- read_rds(file = "data/s_curve/s_curve_pacmap.rds")

#(knn: 10, init: random, \n MN-ratio: 0.9, FP-ratio: 2)
plot_list5 <- PaCMAP_data |>
  ggplot(aes(x = PaCMAP1,
             y = PaCMAP2)) +
  geom_point(alpha=0.1, size=1, colour="#ff7f00") +
  annotate(geom = 'text', label = 'e', x = Inf, y = Inf, hjust = 1.5, vjust = 1.5, size = 3)

```{r fig-nldervis, fig.cap="2D layouts from different NLDR techniques applied the same data: (a) tSNE (perplexity = 27), (b) UMAP (n_neighbors = 50), (c) PHATE (knn = 5), (d) TriMAP (n_inliers = 5, n_outliers = 4, n_random = 3), and (e) PaCMAP (n_neighbors = 10, init = random, MN_ratio = 0.9, FP_ratio = 2). Is there a best representation of the original data or are they all providing  equivalent information?"}

plot_list1 + plot_list2 + plot_list3 + plot_list4 + plot_list5 +
  plot_layout(ncol=5)
```

@@

## Package dependencies

Understanding the dependencies of the "quollr" package is essential for smooth operation and error prevention. The following dependencies refer to the other R packages that `quollr` relies on to execute its functions effectively.

```{r}
package_dependencies("quollr")
```

## Usage
